Can We Crunch the Social Graph on the Cheap?

### 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex
approach changes and comments

And these graphs can be large---too large for a single node \todo{may want to avoid saying 'node' since graph 'node' is also used in this paragraph}. In January 2011, Facebook had 500 million active users with an average of
130 friends each \cite{Facebook:2011p91}. A natural approach would be
to try to partition the graph across the nodes in a cluster. But with
low diameter graphs, a traversal starting at any node is likely to
quickly leave its partition and travel over the network, incurring
high latency costs. And power law graphs have a small number of nodes
that generate significantly more work than other nodes, leading to
computational imbalance.

@@

Each coroutine executes non-preemptively, utilizing the core to which
its ``parent'' thread has been assigned by the operating system until it
reaches a yield point.  Typically, the coroutine yields after
initiating a prefetch operation on data that the programmer or
compiler anticipates would probably not complete before the data is
referenced by a subsequent load operation.  In this way, the core
continues to execute instructions for other coroutines when otherwise
it would likely stall.  The programming model thus tolerates latency.

@@

Our goal is to enable high throughput on graphs with two challenging
properties. First, the graphs' combination of sparseness and low
degree means there is little locality to be exploited. Second, the
graphs are larger than a single node's memory and do not partition
well \todo{explain?}. We must spread the graph across multiple nodes
and send memory operations over a network. If memory requests are
uniformly distributed over the working set, a given node will make as
many accesses to a distant node's memory as it does to the local node.

We would like to run as fast as current hardware will let us. Pin
bandwidth is the tightest resource: Intel's QuickPath Interconnect
(QPI) allows approximately 100 million requests per second
\todo{cite}, and recent interconnect products allow similar request
rates, with a round-trip latency of approximately 3$\mu$s
\todo{cite}. Little's law tells us then that we must support on the
order of 300 concurrent memory operations, and thus coroutines, to
cover this latency.

These requirements drive the design of our system. There are three key
components.

\subsection{Low-latency context switch}

There are three main concerns in designing the coroutine
mechanism. First, contexts must be small, so many coroutines can be
active without monopolizing the cache. Second, context switches must
be fast, so that overhead does not swamp actual work. Third, when a
coroutine yields, we must choose a good coroutine to run next.

@@

The choice of which coroutine to run next presents us with a
tradeoff. The simplest scheduler chooses coroutines to execute in
round-robin fashion. But this may lead to performance problems: a
coroutine might be scheduled before the event which caused it to yield
has completed, stalling the pipeline and keeping other coroutines from
progressing. A coroutine might also be scheduled long after the event
which caused it to yield has completed; while this wouldn't block
other coroutines from proceeding, it would add unnecessary latency to
the yielding coroutine. A more intelligent scheduler might allow
coroutines to execute only when they were ready, but this could
increase the overhead of context switching. Our initial implementation
takes the simple approach, but more investigation is warranted.

@@

In order to minimize memory access overhead, the memory operation
library calls do not interact with the network interface
directly. Instead, they queue their operations to another thread, the
{\em memory manager thread}, running on another core in the CPU. This thread
translates the global address into a network destination and issues
the network operation, checks for completion, and manages the returned
data buffers.

But we have to be careful---directing memory requests from multiple
compute threads to a single memory manager thread may reduce memory
concurrency. And we've increased the overhead of a single memory
operation---what in the single node case was a simple read, is now
multiple queuing operations between threads, and multiple reads and
writes over QPI to the network interface. Adding more memory manager
threads may enable more memory concurrency, but the overhead is still high.

Instead, we propose hardware acceleration with a {\em memory manager
  FPGA}, a coprocessor FPGA inserted into a Xeon socket so it can
participate in the node's coherence domain. This coprocessor does no
computation---it would only manage the global memory requests of the
main CPU, performing the function of the memory manager thread
described earlier.

@@

Unfortunately, current processors do not provide sufficient hardware
resources for us to reach our concurrency target using this
approach. Each core in Intel's X5650 Xeon supports only 10 outstanding
misses from its private L2; the entire chip is limited to 32 L2 misses
to local memory and 12 L2 misses for memory reached over the QPI
bus. A simple translation of a local load to a network access would
tie up one of these slots for the entire multi-microsecond latency of
the request. We must bypass these structures to allow the amount of
concurrency we need; the CPU needs to direct the memory manager FPGA
to do work on its behalf without tying up its own resources.

To accomplish this, we encode commands in the high-order bits of
remote memory addresses. The memory manager FPGA advertises each
remote address twice. The first address is used for blocking reads and
writes---any access incurs the full latency of the remote
access. \todo{verify writes block?} The second address is used for
prefetches and asynchronous writes. These addresses differ by one
high-order bit, easily set with a mask in a macro. A read from the
prefetch address starts the network transaction, but immediately
returns a dummy value. When the value arrives, it is stored in a
temporary buffer; when the requesting coroutine is scheduled again, it
executes the blocking read, and the memory manager immediately returns
the value from the buffer. \todo{this isn't fully baked---do we want
  to include this sort of detail?}

@@

In systems with hundreds of threads per node, lightweight
synchronization is important. While we have not implemented any
synchronization primitives yet, full-empty bits are a natural choice
given our programming model. Two questions then arise: at what
granularity should full-empty bits work, and what component should do
the synchronization?

@@

\section{Current Implementation and Results}

\todo{Begin eval insert}
\input{evaluation}
\todo{End eval insert}

@@

why not wait for more cores?

>>>>

And these graphs can be large---too large for a single node. In January 2011, Facebook had 500 million active users with an average of
130 friends each \cite{Facebook:2011p91}. A natural approach would be
to try to partition the graph across the nodes in a cluster. But with
low diameter graphs, a traversal starting at any vertex is likely to
quickly leave its partition and travel over the network, incurring
high latency costs. And power law graphs have a small number of vertices
that generate significantly more work than other vertices, leading to
computational imbalance.

@@

Each coroutine executes non-preemptively \todo{seems like non-premptively isn't ``how" it executes, rather how it is scheduled, maybe say ``cooperatively"?}, utilizing the core to which
its ``parent'' thread has been assigned by the operating system until it
reaches a yield point.  Typically, the coroutine yields after
initiating a prefetch operation on data that the programmer or
compiler anticipates would probably not complete before the data is
referenced by a subsequent load operation.  In this way, the core
continues to execute instructions for other coroutines when otherwise
it would likely stall.  The programming model thus tolerates latency.

@@

Our goal is to enable high throughput on graphs having two challenging
properties. First, the graphs' combination of sparseness and low
degree means there is little locality to be exploited. Second, the
graphs are larger than a single node's memory and do not partition
well because of low diameter \todo{enough explanation now?}. We must spread the graph across multiple nodes
and send memory operations over a network. If memory requests are
uniformly distributed over the working set, a given node will make as
many accesses to remote nodes' memory as there are remote requests to the local node's memory coming in.

We would like to run as fast as current hardware will let us. Pin
bandwidth is the tightest resource: Intel's QuickPath Interconnect
(QPI) allows approximately 100 million requests per second
\todo{cite}, and recent interconnect products allow similar request
rates, with a round-trip latency of approximately 3$\mu$s
\todo{cite}. Little's law tells us then that we must support on the
order of 300 concurrent memory operations--and thus, up to this many coroutines--to
cover this latency.

These requirements drive the design of our system. There are three key
components. First, the system needs lightweight context switches (coroutines) to be able to support many contexts--thereby increasing memory concurrency available--with minimal impact on performance. Second, a memory management subsystem must support many concurrent accesses to a global address space. Third, the system must provide synchronization at a fine-granularity \todo{this last one sounds incomplete because no reason given}.

\subsection{Lightweight context switch}
\todo{can we include in the title something that means lightweight and good scheduling?}

There are three main concerns in designing the coroutine
mechanism. First, contexts must be small, so many coroutines can be
active without monopolizing the cache. Second, context switches must
be fast, so that overhead does not swamp actual work. Third, when a
coroutine yields, we must choose a suitable coroutine to run next.

@@

The choice of which coroutine to run next presents us with a
tradeoff. The simplest scheduler chooses coroutines to execute in
round-robin fashion. But this may lead to performance problems: a
coroutine might be scheduled before the event which caused it to yield (e.g., a long latency memory access)
has completed, stalling the pipeline and keeping other coroutines from
progressing. A coroutine might also be scheduled long after the event
which caused it to yield has completed; while this wouldn't block
other coroutines from proceeding, it would add unnecessary latency to
the yielding coroutine. A more intelligent scheduler might allow
coroutines to execute only when they were ready, but this could
increase the overhead of context switching. Our initial implementation
takes the simple approach, but more investigation is warranted.

@@

\subsubsection{Software management}
In order to minimize memory access overhead, the memory operation
library calls do not interact with the network interface
directly. Instead, they queue their operations to another thread, the
{\em memory manager thread}, running on another core in the CPU. This thread
translates the global address into a network destination and issues
the network operation, checks for completion, and manages the returned
data buffers.

But we have to be careful---directing memory requests from multiple
compute threads to a single memory manager thread may reduce memory
concurrency. And we have increased the overhead of a single memory
operation: what in the single node case was a simple read, is now
multiple queuing operations between threads and multiple reads and
writes over QPI to the network interface. Adding more memory manager
threads may enable more memory concurrency, but the overhead is still high \todo{why? because lots of code to run? so accel by impl in hardware. Maybe this is answered by the "what in...is now" already}

\subsubsection{Hardware management}
\todo{might want to call it accelerator throughout after saying FPGA once}
Considering this overhead, we propose hardware acceleration with a {\em memory manager
  FPGA}, a coprocessor FPGA inserted into a Xeon socket so it can
participate in the node's coherence domain. This coprocessor does no
computation---it would only manage the global memory requests of the
main CPU, performing the function of the memory manager thread
described earlier.

@@

Unfortunately, current processors do not provide sufficient hardware
resources for us to reach our concurrency target using this
approach. Each core in Intel's X5650 Xeon supports only 10 outstanding
misses from its private L2; the entire chip is limited to 32 L2 misses
to local memory and 12 L2 misses for memory reached over the QPI
bus. A simple translation of a local load to a network access would
tie up one of these slots for the entire multi-microsecond latency of
the request. To allow the amount of concurrency we need, we must bypass these structures; the CPU needs to direct the memory manager FPGA
to do work on its behalf without tying up its own resources.

To accomplish this, we encode commands in the high-order bits of
remote memory addresses. The memory manager FPGA advertises each
remote address twice \todo{this sounded like an active thing the FPGA is doing each time you make remote request, first time I read it. Consider just sticking to cmd as upper order address bits than saying advertise each address twice}. The first address is used for blocking reads and
writes---any access incurs the full latency of the remote
access. \todo{verify writes block?} The second address is used for
prefetches and asynchronous writes. These addresses differ by one
high-order bit, easily set with a mask in a macro. A read from the
prefetch address starts the network transaction, but immediately
returns a dummy value. When the value arrives, it is stored in a
temporary buffer; when the requesting coroutine is scheduled again, it
executes the blocking read, and the memory manager immediately returns
the value from the buffer \todo{is this necessarily restricted to the memory manager returning value from buffer rather than buffer in local memory?  maybe it is, since if buffer was in local mem, would have to also write a full bit or make two reads: valid bit and data}. \todo{this isn't fully baked---do we want
  to include this sort of detail?}

@@

In systems with hundreds of threads per node, lightweight
synchronization is important. While we have not implemented any
synchronization primitives yet, full-empty bits are a natural choice
given our programming model \todo{say why natural choice unless programming model section makes this point already}. Two questions then arise: at what
granularity should full-empty bits work, and what component should do
the synchronization?

@@

% 2TB global memory => 68GB of distributed FE bits.  This seems reasonable, given that a 2TB system might have 64 nodes, so ~1GB per memory manager if one manager per node.

\section{Current Implementation and Results}
\input{evaluation}

@@

why not wait for more cores?
- perhaps memory subsystems on a chip are not keeping up with the bandwidth of so many cores.
